### 2025-10-31 write-ups/manuscript/kid-tangrams-ms.Rmd
intro and discussion

authornote: |
  TODO credit taxonomy etc

abstract: > 
  Learning language requires learning not only the content of language, but also how to use language to communicate. Iterated reference games provide a window into such skills, requiring rich communication as participants converge on mutually understandable names for initially novel referents. Some early experiments are interpreted as evidence that 4-5-year-old children cannot generate the mutually understandable names needed to succeed in an iterated reference game. Here, we revisit young children's referential communicative abilities using a simpler, child-friendly paradigm. Across TODO pairs of children, we found that 4-5-year-olds successfully established reference with each other. Children were TODO% accurate, and they often used descriptions similar to their partner's. These findings suggest that children’s capacity to construct effective referring expressions in novel contexts emerges earlier than once thought, consistent with the view that children show early pragmatic competence in supportive contexts.
  <!-- https://tinyurl.com/ybremelq -->

@@

TODO review the literature for 
* any other studies we should include
* any studies we want to say more about than we had room for in cogsci

@@

TODO this description metric is just the length of the description, taking out repetition/rephrasing from the child. We could also include the robustness check which is the length of all the (on-topic) teller-guesser interaction (including repeating/rephrasing, questions, I don't knows, etc)

TOOD maybe move this paragraph Some examples to illustrate the variety of effective descriptions children employed are shown in Table \ref{example} (these specific examples are from Experiment 2, but both experiments had similar distributions of descriptions). TODO add examples from expt 3 as well??

@@

We compared the semantic similarities between descriptions of the same target based on who produced the description (Figure \ref{sim}).  We used a Bayesian mixed effects linear regression to predict similarity.^[`r sim_across_between_form |> form()`] Utterances were more similar if they came from the same partnership (`r sim_across_between_1_results |> stats_text(2,3)`). Utterances were slightly more similar still if they came from the same person (`r sim_across_between_1_results |> stats_text(3,3)`), which is expected since children are likely to be fairly consistent with themselves. However, children used descriptions that were much more similar to their partner's than to other children's (Figure \ref{sim}), indicating sensitivity to their partner's expressions. 
TODO consider dropping this because it's not that strong an analysis?

@@

### Participants
Experiment 3 was run between XX and YY , at the same preschool as Experiments 1 and 2. Because the stimuli changed, we allowed children who had participated in a prior experiment to participate again, and ZZ participants in experiment 2 were also in experiment 3. TODO pairs of children completed all 16 critical trials, and TODO. Our target age range was 4 and 5-year-olds. Of the TODO children, QQ were girls, and the children had a median age of ZZ months, and a range of XX-YY months. An additional XX games were excluded for being pilot runs, including repeat participants, not completing enough trials, or including a 3-year-old participant.

@@

### Data processing
Data were processed in the same way as Experiment 1. After excluding trials where children did not give a description or where the experimenter echoed a child's description, we had TODO trials total.

@@

TODO reverify everything after data review + full sample

@@

We repeat the semantic analyses from Experiment 2. To start with, we again found that utterances were more similar if they came from the same partnership (`r sim_across_between_3_results |> stats_text(2,3)`) and if they came from the same person (`r sim_across_between_3_results |> stats_text(3,3)`).

Over the first three blocks, descriptions became increasingly similar to the last block description (`r sim_to_last_3_results |> stats_text(2,3)`). Over time, the semantic distance between adjacent block utterances increased slightly (`r sim_to_next_2_results |> stats_text(2,3)`).

@@

As the two experiments were similar to one another, we re-ran models pooling the data across the three experiments, using experiment number as a random effect. Pooling the three experiments, children's accuracy was above chance (OR: `r stats_text(acc_mod_meta,1)`) and accuracy numerically increased over the course of the game, although the credible interval included 0 (OR of one trial later: `r stats_text(acc_mod_meta, 2)`).

TODO decide whether to discuss speed here again.

@@

For other semantic analyses, we pooled across experiments 2 and 3 which had 4 blocks. In the combined model, we found that the similarity the last block utterance increased across blocks (`r stats_text(sim_to_last_mod_meta, 2,3)`) and was higher when the same child produced both utterances (`r stats_text(sim_to_last_mod_meta, 3,3)`). In a combined model, similarity to the next block utterance also increased across blocks (`r stats_text(sim_to_next_mod_meta, 2,3)`) and was higher when the same child produced both utterances (`r stats_text(sim_to_next_mod_meta, 3,3)`). Across the two experiments, utterances from different pairs to the same stimulus tended to decrease in similarity across blocks (`r stats_text(sim_across_mod_meta, 2,3)`).

While some of these effects were small and thus not all effects were robustness detectable in each experiment by itself, the overall picture is consistent in direction with the patterns of convergence and divergence reported in reference games between adults [@boyce2024].

@@

We did not account for children's age and gender when pairing children to play the game, so children of different ages (within the 4-5 year old target age range) sometimes played together, and children sometimes played with same-gender and sometimes with different-gender partners. As children of this age are rapidly developing their linguistic skills (TODO CITATION NEEDED), we looked at possible effects of children's age and gender on performance as an exploratory measure.

@@

The effects of children's age and gender on their performance are shown in Figure \ref{ageacc}. We ran a mixed effects logistic regression, where we considered possible effects of the age of each child (centered at mean age of all the children who participated) and the gender of each child^[`r form(acc_demog_form)`]. Accuracy was higher for older children (OR for one month older speaker: `r stats_text(acc_demog_meta, 3,2)`, OR for one month older listener: `r stats_text(acc_demog_meta, 2,2)`). Accuracy was also higher when the listener was a girl (OR: `r stats_text(acc_demog_meta, 4,2)`), but was not credibly higher when the speaker was a girl (OR: `r stats_text(acc_demog_meta, 5,2)`).

For description length, we tested whether children's age or gender effected how many words they produced in descriptions when in the teller role (Figure \ref{agewords}). We also ran a mixed effects linear regression, where we considered the ages and genders of both children as possible predictors^[`r form(description_demog_form)`]. Older children tended to produce longer descriptions (effect of one month older `r stats_text(description_demog_meta,3,2)`), but there was no effect of age of listener (`r stats_text(description_demog_meta, 2,2)`). For gender, there was no effect of teller gender on description length (girls produced `r stats_text(description_demog_meta, 5,2)` words compared to boys), but tellers produced longer descriptions for boy guessers (`r stats_text(description_demog_meta, 4,2)`). There were not substantial effects of trial order or interactions between demographics and trial order.

It is curious that listener gender but not speaker gender affected description length, but for this task, it is unclear how description length would be expected to map onto linguistic sophistication for this task and demographic. For accuracy, the age and gender effects could be reflective of any of a number of developing capacities, including linguistic, executive functioning, or social/theory of mind. 
These findings of a slight effect of age and a slight female-advantage are consistent with a wider literature of TODO

@@

# (unclear if include)
could also do some qualitative comparison of the *types* of descriptions that kiddos give versus kid-parent vs adult ?

@@

Across 3 experiments and TODO pairs of 4 and 5-year-old children, we tested how well children could produce referential expressions that allowed their partner to find a matching abstract shape. Children varied substantially in what sorts of descriptions they produced, but overall accuracy was high (TODO%), indicating that children were generally able to produce adequate descriptions. Additionally, children's utterances showed signs of converging toward conceptual pacts. While this task is substantially scaled down relative to measures used for adult competence, it does suggest that the relevant communication skills are present at least in rudimentary form by the end of the preschool years.

@@

The generalizability of our results is limited by the target population, the target images, and the task structure. We sampled a convenience population of children at a university nursery school. The set of tangram images may be easier to distinguish and have higher codability than other target images used in adult reference games. We specifically targeted children's abilities to construct referring expressions that can be jointly understood, so children were provided scaffolding around taking turns and talking to their partner. Thus, children's performance should be taken as a proof-of-concept about ability, rather than a claim about how generally children spontaneously demonstrate these abilities.

Even within this population, we observed some substantial pair to pair differences. Pairs varied in what types of descriptions they have, how readily the gave them, as well as how well they understood the "rules" of the game and how much they cooperated with the rules. While we did not aim to test 3 year olds, the few 3 year olds who we tested accidentally did fine.

Our qualitative impression is that it was more common for children to hit limits around understanding the game structure (ex. that the "teller" doesn't press the ipad),  cooperating with the game (ex. not trying to show the ipad to their partner, not pressing buttons before their partner has a change to describe), and understanding that this wasn't a test about the "name" of the shape, moreso than being truely able to describe an unknown referent adequately.
Pairs of children did very more in how sensitive they were to collaborating with their partner and adapting descriptions based on their partners descriptions and their prior success. Generally, it's hard to push at the lower limits of some of children's abilities because many skills are changing over the same time frame and variables that may not be of primary interest, such as patience and instruction-following, can influence the outcomes.

>>>>

authornote: |
  TODO credit taxonomy, acknowledgements, etc

abstract: > 
  Learning language requires learning not only the content of language, but also how to use language to communicate. Iterated reference games provide a window into such skills, requiring rich communication as participants converge on mutually understandable names for initially novel referents. Some early experiments are interpreted as evidence that 4-5-year-old children cannot generate the mutually understandable names needed to succeed in an iterated reference game. Here, we revisit young children's referential communicative abilities using a simpler, child-friendly paradigm. Across TODO pairs of children, we found that 4-5-year-olds successfully established reference with each other. Children were TODO% accurate, and they often used descriptions similar to their partner's. These findings suggest that children’s capacity to construct effective referring expressions in novel contexts emerges earlier than once thought, consistent with the view that children show early pragmatic competence in supportive contexts.

@@

@@

TODO this description metric is just the length of the description, taking out repetition/rephrasing from the child. We could also include the robustness check which is the length of all the (on-topic) teller-guesser interaction (including repeating/rephrasing, questions, I don't knows, etc), which is also very flat.

[TODO maybe move this paragraph] Some examples to illustrate the variety of effective descriptions children employed are shown in Table \ref{example} (these specific examples are from Experiment 2, but both experiments had similar distributions of descriptions).

@@

We compared the semantic similarities between descriptions of the same target based on who produced the description (Figure \ref{sim}).  We used a Bayesian mixed effects linear regression to predict similarity.^[`r sim_across_between_form |> form()`] Utterances were more similar if they came from the same partnership (`r sim_across_between_1_results |> stats_text(2,3)`). Utterances were slightly more similar still if they came from the same person (`r sim_across_between_1_results |> stats_text(3,3)`), which is expected since children are likely to be fairly consistent with themselves. However, children used descriptions that were much more similar to their partner's than to other children's (Figure \ref{sim}), indicating sensitivity to their partner's expressions.

TODO consider dropping this because it's not that strong an analysis?

@@

### Participants
Experiment 3 was run between XX and YY, at the same preschool as Experiments 1 and 2. Because the stimuli changed, we allowed children who had participated in a prior experiment to participate again, and ZZ participants in experiment 2 were also in experiment 3. TODO pairs of children completed all 16 critical trials, and TODO others completed ... . Our target age range was 4 and 5-year-olds. Of the TODO children, QQ were girls, and the children had a median age of ZZ months, and a range of XX-YY months. An additional XX games were excluded for being pilot runs, including repeat participants, not completing enough trials, or including a 3-year-old participant.

@@

### Data processing
Data were processed in the same way as Experiment 1. After excluding trials where children did not give a description or where the experimenter echoed a child's description, we had `r joint_data |> filter(expt=="3") |> nrow()` trials total.

@@

TODO reverify everything after full sample

@@

We repeated the semantic analyses from Experiment 2. We again found that utterances were more similar if they came from the same partnership (`r sim_across_between_3_results |> stats_text(2,3)`) and if they came from the same person (`r sim_across_between_3_results |> stats_text(3,3)`).

Over the first three blocks, descriptions became increasingly similar to the last block description (`r sim_to_last_3_results |> stats_text(2,3)`). Over time, the semantic similarity between adjacent block utterances increased slightly (`r sim_to_next_2_results |> stats_text(2,3)`).

@@

As the experiments were all similar to one another, we re-ran models pooling the data across the three experiments, with experiment number as a random effect. Pooling across the three experiments, children's accuracy was above chance (OR: `r stats_text(acc_mod_meta,1)`) and accuracy numerically increased over the course of the game, although the credible interval included 0 (OR of one trial later: `r stats_text(acc_mod_meta, 2)`).

TODO decide whether to include speed here again.

@@

For the other semantic analyses, we pooled across experiments 2 and 3 which had 4 blocks. In the combined model, we found that the similarity the last block utterance increased across blocks (`r stats_text(sim_to_last_mod_meta, 2,3)`) and was higher when the same child produced both utterances (`r stats_text(sim_to_last_mod_meta, 3,3)`). In a combined model, similarity to the next block utterance also increased across blocks (`r stats_text(sim_to_next_mod_meta, 2,3)`) and was higher when the same child produced both utterances (`r stats_text(sim_to_next_mod_meta, 3,3)`). Across the two experiments, utterances from different pairs to the same stimulus tended to decrease in similarity across blocks (`r stats_text(sim_across_mod_meta, 2,3)`).

While some of these effects were small and thus not all effects were robustly detectable in every experiment, the overall pattern of results is consistent in direction with the patterns of convergence and divergence reported in reference games between adults [@boyce2024].

@@

Children in the late preschooler years are undergoing a time of rapidly developing linguistic and cognitive skills, so children may show age-based differences in how well they complete tasks that use language or executive functioning [citation needed]. Working memory and other executive capacities are developing across childhood [@diamond2013]. Additionally, there may be gender differences at this age. Young girls have on average an advantage in early linguistic skills, showing faster earlier vocabulary growth that boys [@marchman; @rinaldi2023], and this linguistic advantage continues until .... TODO

We did not design these experiments to test for the affect of children's age or gender. Children were not matched with their partner based on age and gender, so different ages (within the 4-5 year old target age range) sometimes played together, and children sometimes played with same-gender and sometimes with different-gender partners. We consider the effects of participant age and gender as an exploratory measure.

@@

The effects of children's age and gender on their performance are shown in Figure \ref{ageacc}. We ran a mixed effects logistic regression, where we considered possible effects of the age of each child (centered at mean age of all the children who participated) and the gender of each child on the accuracy of their selections^[`r form(acc_demog_form)`]. Accuracy was higher for older children (OR for one month older speaker: `r stats_text(acc_demog_meta, 3,2)`, OR for one month older listener: `r stats_text(acc_demog_meta, 2,2)`). Accuracy was also higher when the listener was a girl (OR: `r stats_text(acc_demog_meta, 4,2)`), but was not credibly higher when the speaker was a girl (OR: `r stats_text(acc_demog_meta, 5,2)`).

TODO be consistent about teller/guesser versus speaker/listener

We also tested whether children's age or gender affected how many words they produced in their descriptions (Figure \ref{agewords}). We ran a mixed effects linear regression, where we considered the ages and genders of both children as possible predictors of description length^[`r form(description_demog_form)`]. Older children tended to produce longer descriptions (effect of one month older `r stats_text(description_demog_meta,3,2)`), but there was no effect of age of listener (`r stats_text(description_demog_meta, 2,2)`). For gender, there was no effect of teller gender on description length (girls produced `r stats_text(description_demog_meta, 5,2)` words compared to boys), but tellers produced longer descriptions for boy guessers (`r stats_text(description_demog_meta, 4,2)`). There were not substantial effects of trial order or interactions between demographics and trial order.

It is curious that listener gender but not speaker gender affected description length, but for this task, it is unclear how description length would be expected to map onto linguistic sophistication for this task and demographic. At one end, longer descriptions may provide more information than short non-specific labels such as "person" or "diamond". On the other, succinct descriptions that reflect distinguishing features ("mountain", "man with backpack") are better than long descriptions about the component shapes ("there's a triangle and then a square and another square...").

For accuracy, the age and gender effects could be reflective of any of a number of developing capacities, including linguistic sophistication, improving executive functioning and impulse control, and better perspective taking and theory of mind.

@@

@@

Across 3 experiments and TODO pairs of 4 and 5-year-old children, we tested how well children could produce referential expressions that allowed their partner to find a matching abstract shape. Children varied substantially in what sorts of descriptions they produced, but overall accuracy was high (TODO%), indicating that children were generally able to produce adequate descriptions. Additionally, children's utterances showed signs of converging toward conceptual pacts. While this task was substantially scaled down relative to measures used for adult competence, it does suggest that the relevant communication skills are present at least in rudimentary form by the end of the preschool years.

@@

The generalizability of our results is limited by the target population, the target images, and the task structure. We sampled a convenience population of children at a university nursery school. We used two sets of 4 tangram images, which may be easier to distinguish and have higher codability than other target images used in adult reference games. We specifically targeted children's abilities to construct referring expressions that can be jointly understood, so children were provided scaffolding around taking turns and talking to their partner. Thus, children's performance should be taken as a proof-of-concept about ability, rather than a claim about how generally children spontaneously demonstrate these abilities.

Even within this population, we observed some substantial pair to pair differences. Pairs varied in what types of descriptions used, how readily they produced them, as well as how well they understood the "rules" of the game and how much they cooperated with the rules.

Our qualitative impression is that it was more common for children to hit limits around understanding the game structure (ex. that the "teller" doesn't press the iPad),  cooperating with the game (ex. not turning their iPad to show their partner, not pressing buttons before their partner has a chance to describe), and understanding that this wasn't a test about the "name" of the shape, than for children to specifically struggle with how to describe an unknown referent adequately.
Pairs of children did vary more in how sensitive they were to collaborating with their partner and adapting descriptions based on their partners descriptions and their prior success. Generally, it's hard to push at the lower limits of some of children's abilities because many skills are changing over the same time frame and variables that may not be of primary interest, such as patience and instruction-following, can influence the outcomes. [TODO do we want to explicitly comment that we aren't making claims about what the lower limit is]
